A Recipe for Scavenging - the natural history of a behaviour

### 2016-04-20 natural history of scavenging.tex
Sorry about the hold up

%%KH Is there anything we should add about other extinct groups such as plesiosaurs and https://en.wikipedia.org/wiki/Plesiosauria#Gait

@@

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move, which may be due to the low muscular efficiency of running \citep{tucker1975energetic} as well as the relative inefficiency of gas exchange in the case of terrestrial mammals \citep[cf. birds and fish ][]{williams1999evolution}. %TG: Hmm I get the point of gas exchange efficiency but (1) it should be more developed if left like that (i.e. what the flip is meant by gas efficient exchanges) and (2) that works for mammals, but in terms of terrestrial that ignores birds, squamates, crocodiles and amphibians.
%
%TG: It's missing a transition here, we go from gas exchanges to posture. Maybe we need something like "Among terrestrial scavengers, most are mammals. This might be due to the posture blablabalba." But then it doesn't connect logically with the sentence before: "mammals have are inefficient...". Maybe one way is to just remove the gas exchange and start with something like:
%Terrestrial environments are the most energetically costly in which to move, which may be due to the low muscular efficiency of running \citep{tucker1975energetic}.
%Most terrestrial species that have been observed (or are suspected to be) scavenging are mammals.
%TG: and then link to the posture hypothesis (with the "meh, we're not sure" at the end).
%
However, the evolutionary transition in posture from the sprawling gait of reptiles to the erect posture of mammals has often been supposed as conferring a huge advantage to the latter.
The purported advantages include benefits in terms of speed, efficiency, muscle effort and manoeuvrability \citep{sullivan2015posture}. 
Despite being intuitive, \cite{sullivan2015posture} states most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
Metabolic rate however, unquestionably impacts terrestrial species whereby ectotherms such as many modern reptiles, cannot move for sustained periods \citep{bennett1979endothermy}. 
% Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}. 
This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
This would also have been true of extinct species with the same physiology. 
%TG: I guess this is a comment too: Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
%Both these groups are more likely to depend on low energetic costs in order to compensate for low encounter rates due to their inability to move. 
It is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion. 
Although the earliest evidence of vertebrate scavenging comes from the Permian (298.9 - 252.17 Mya) where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
%TG: I do like the way all our paragraphs end up with a "But wait, there's an exception!"

@@

Of course, terrestrial animals can also move bipedally. 
%TG: missing some transition here. Maybe something short like "Further back in time" or "Previous to that"...
Although the evolution of bipedal movement was significant in that it freed up the forelimbs for other purposes (e.g. climbing, tool-use, wing development etc.) it does not differ radically in cost from quadrupedal locomotion \citep[][and references therein]{williams1999evolution}. 
For instance, \cite{alexander2004bipedal} shows that, in the case of humans, we are more economical than predicted while walking and less so while running according to predicted costs of terrestrial movement.

Aside from humans and our allies, the best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous theropods.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic \citep[i.e. intermediate to ecto- and endotherms;][]{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}.
Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles.

%TG: Dudes! Cite your godamn Am Nat paper in this paragraph! It's totally more relevant than Weishampel's book!

% KH Something about humans (also the carnivours kangaroo which might be the most efficient mover.)

\subsection*{Detection}
It would be pointless to have incredible ranging abilities and not have the sensory architecture to benefit from it.
%If we came at this from a position of complete ignorance we would predict scavengers to have well-developed senses and indeed, this is what we find. %TG: I don't really like this sentence. How about:
As predicted by the necessity of an increased encounter rate, scavengers have well-developed senses.
A simplification of terrestrial, vertebrate scavengers in sensory terms is one of them existing in a two-dimensional plane while foraging for carrion directly.
They can detect carcasses at a range that is defined by the radius of their sensory organs. %, usually the visual and olfactory senses.
As a consequence, they have a much more restricted view of the landscape than do aerial foragers.
Hyenas make up for this in their ability to smell a rotting carcass 4 km away and to hear the vocalisations of conspecifics at a distance of 10 km \citep{mills1989comparative}. 
% Using the approach of \cite{spiegel2013factors} we estimate a spotted hyena %TG: do we need a (\textit{Crocuta crocuta}) here?
%could resolve a 2 metre target at 1 km distance. 
We can compare this to the energetics approach of \cite{ruxton2004obligate}, who calculated a terrestrial scavenger needs to be able to detect carrion at 500 meters in order to survive, which is clearly within the ability of hyenas.  
% While considering prehistoric habitats \cite{ruxton2004obligate} calculated that ``a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m".
Moreover, the senses of many extant (and in all probability extinct) carnivores meet this required distance, making scavenging feasible for terrestrial species \citep{farlow1994speculations,mech2010wolves}.

Species capable of flight have effectively added an extra spatial dimension (i.e. the vertical component) to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Certainly, vultures are known to have impressive visual acuity, with one estimate indicating lappet-faced vultures %TG: We need a standardisation for english species names, I'll go for all lower case (similar as what you'd do for "pig" or other english common species names).
 (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
Eagles too are known to have highly developed vision \citep{reymond1985spatial}.
It follows that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts \citep{AR:AR22815}.
%We can contrast this ability with bats, whose visual acuity is famously poor. %TG: Humph! That's not nice! I'm wouldn't go with all the "this is known" thingies here. It really lowers down the quality of our claim and is totally un-necessary (and actually probably wrong: lot's of fruit bats (like flying foxes) do not echolocate and thus have to find their fruits among the flipping tropical folliage using their eyes and noses only)...
We can contrast this with other flying vertebrates such as many bats whose reliance on echolocation would not lend itself to discovering immobile carrion.
%

@@

%Thus, the bone crushing ability of hyenas and others reveals another useful scavenger trait. %TG: how about the following for a smoother transition:
In the ability to eat bone scavengers have arrived at a way to feed on a resource that is typically too hard for many predators to process. 
% Thus, some phenotypic adaptations such as the ability to crush bone reveal another useful scavenger trait.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the estimated 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

@@

Interestingly, such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{ARCM:ARCM12084}.
The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years \citep{dominguez2002hunting}.
A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}.
In some historical hominin-inhabited areas there were a greater number of felids than hyenids.
Again, this is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth, enabling contemporaneous hominins to benefit \citep{pobiner2015new}.
The use of tools and the cooperative nature of hominins meant they could likely get a substantial part of their energetic requirements through scavenging depending on their environment \citep{moleon2014humans}.

In Mesozoic systems some extremely large theropod dinosaurs had a morphology indicative of an ability to process bone \citep[e.g. the robust skull and dentition of \textit{Tyrannosaurus rex}][]{hone2010feeding}.
There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}.
The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating} which would have been powerful enough to break open skeletal material \citep{rayfield2001cranial}.

>>>>

%%KH Is there anything we should add about other extinct groups such as plesiosaurs and https://en.wikipedia.org/wiki/Plesiosauria#Gait maybe an air breathing non-airbreathing devide would be good.

@@

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move \citep{tucker1975energetic}. Unlike aerial and aquatic environments support must be provided through the animals posture leading to trade offs between low cost of maintain and the efficiency of transport. For example, the sprawling posture of many ectotherm groups (except frogs!) would be expected to allow for a lower maintenance cost relating to body support in comparison to groups with the limbs position underneath the body, such as mammals, aves and dinosaurs.
%KH: I agree with TG and have removed the gas exchange thing as its a physiology thing only applicable to certain groups and not got to do with terrestrial environments really. I also removed the 
%Also that reference is from 1975, is this up to date?

%TG: Hmm I get the point of gas exchange efficiency but (1) it should be more developed if left like that (i.e. what the flip is meant by gas efficient exchanges) and (2) that works for mammals, but in terms of terrestrial that ignores birds, squamates, crocodiles and amphibians.

%
%TG: It's missing a transition here, we go from gas exchanges to posture. Maybe we need something like "Among terrestrial scavengers, most are mammals. This might be due to the posture blablabalba." But then it doesn't connect logically with the sentence before: "mammals have are inefficient...". Maybe one way is to just remove the gas exchange and start with something like:
%Terrestrial environments are the most energetically costly in which to move, which may be due to the low muscular efficiency of running \citep{tucker1975energetic}.
%Most terrestrial species that have been observed (or are suspected to be) scavenging are mammals.
%TG: and then link to the posture hypothesis (with the "meh, we're not sure" at the end).
%
While the low maintenance costs of a sprawling like posture may reduce the need to scavenge frequently, the ability to search large areas is several restricted in these groups hence likely reducing such opportunities (ref on home range of terrestrial ectoterms). For example, a sprawling gait can results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. On the other hand the more erect body positions and often associated higher basal metabolic rates, allow for sustained locomotion over long periods of time and hence allow such species to search large areas.

The early transition from a sprawling gait, such as seen in Tiktaalik, to the more erect posture of Synapsids and later dinosaurs and mammals, has often been supposed as conferring a huge advantage to the latter.
%KH Even after changing it I dont like this transition type narrative.
%KH also needs to be more clear on what the advantage to scavenging is for these groups.
%The purported advantages include benefits in terms of speed, efficiency, muscle effort and manoeuvrability \citep{sullivan2015posture}. 
%none of these advantages refer very clearly to scavenging
Despite being intuitive, \cite{sullivan2015posture} states most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
%KH I'm always wary of deeming something as intuitive,

%Metabolic rate however, unquestionably impacts terrestrial species whereby ectotherms such as many modern reptiles, cannot move for sustained periods \citep{bennett1979endothermy}. 
% Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}.

%KH I think as ectothermy is almost completely associated with sprawling gaits I would tie it up together to avoid repeating something similar here.

%This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
%This would also have been true of extinct species with the same physiology. 
%TG: I guess this is a comment too: Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
%Both these groups are more likely to depend on low energetic costs in order to compensate for low encounter rates due to their inability to move. 
Whatever the case, it is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion. 
Although the earliest evidence of vertebrate scavenging comes from the Permian (298.9 - 252.17 Mya) where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
%TG: I do like the way all our paragraphs end up with a "But wait, there's an exception!"

@@

While we can use modern analogues of quadrupedal species, the absence of scavenging in bipedal animals makes this more difficult. However, as a groups bipedal locomotion may have been a particularly useful strategy for  searching large areas. For instance, \cite{alexander2004bipedal} shows that, in the case of humans, we are more economical than predicted while walking according to predicted costs of terrestrial movement. In fact the ability of humans to engage in hunting strategies that involve traversing long ranges, such as persistence hunting, has led to several researches highlighting the possible role of scavenging in early hominids. Indeed the low energetics of human walking at least allows for a substantial amount of scavenging in our hominid ancestors (Graemes papers on scavenging).
%%KH I know this is excluding other hominids so feel free to write a more inclusive line.

%Of course, terrestrial animals can also move bipedally. 
%TG: missing some transition here. Maybe something short like "Further back in time" or "Previous to that"...
%I agree, 
%Although the evolution of bipedal movement was significant in that it freed up the forelimbs for other purposes (e.g. climbing, tool-use, wing development etc.) it does not differ radically in cost from quadrupedal locomotion \citep[][and references therein]{williams1999evolution}. 
%For instance, \cite{alexander2004bipedal} shows that, in the case of humans, we are more economical than predicted while walking and less so while running according to predicted costs of terrestrial movement. 
%the second sentence does not really follow from the first -
Aside from humans and our allies, the best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been extensively explored in the carnivorous theropods.
%KH you could reference all the shit on theropod scavenging here if you wanted.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
While the locomotory ability of theropods has been debate since their first inception, more recent studies have reconstructed them as relatively mobile animals (John Hutchinsons papers). 
%See that Pontzer 
Despite some suggestions that the large size of some forms may have allowed some advantage in scavenging, partially due to the abilty to search large areas (Graemes paper etc), more recent studies have shown that the energetic demands of such large forms meant that scavenging was likely to be more prevalent in mid sized theropods of approximately half and thon (Us and Carbones paper).

%They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
%Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic \citep[i.e. intermediate to ecto- and endotherms;][]{grady2014evidence}.
%KH Do we need to get into this?
%We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}.
%KH Is this not a bit trivial almost?
%Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles. 
%I don't think we can say that, we have no idea on thier range. For example using body size they should be larger than modern mammals, however the importance of metabolic rate is still not clear with home range (body size explains most.). This might be useful http://www.jstor.org/stable/10.1086/682070?seq=1#page_scan_tab_contents

%TG: Dudes! Cite your godamn Am Nat paper in this paragraph! It's totally more relevant than Weishampel's book!
%KH I think the terrestrial section needs a little more work still, it doesn't hang together quite right yet although it is not far off it. While I have wrote it up this way I think ending with humans would be good as it gives a returning to the present type feel which wraps the whole section up without being too explicit about it. Maybe using large bird locomotion would allow this.

\subsection*{Detection}

%KH Just my opinion but I found it really difficult to understand the flow of this section. I starts off with a type of 2d- 3d thing but then kinda bounces around different ideas with everything talked about in relation to aerial viewing. Its also missing how understanding detection ranges can help us understand past scavenging behavior. I would include in the terrestrial section, the use of looking at the portion of the brain attributed to olfactory sensing (like the way we did the scaling of detection range in our paper) as a possible indicator of scavenging. I would also throw into the social communication section that these behaviors can show how groups can be good scavengers through active/passive communication and we might expect that in other extinct social groups (the vulture thing was almost impossible for pterodactyls to avoid I would say and probably common in many soaring flying species). Also Graeme has a paper on human scavenging about how hominies may have exploited vultures and also probably searched in a social sense. 
%I might be too grumpy at the end of the day, but I think a bit of a restructure might help this section.

%It would be pointless to have incredible ranging abilities and not have the sensory architecture to benefit from it.
%KH I also don't like this sentence as it wouldn't be pointless, 
%If we came at this from a position of complete ignorance we would predict scavengers to have well-developed senses and indeed, this is what we find. %TG: I don't really like this sentence. How about:
As predicted by the necessity of an increased encounter rate, scavengers have well-developed senses, with the visual and olfactory sensory systems most often associated with scavenging behavior. This is perhaps no surprise as the nature of carrion means that sensory systems that rely on detecting signals associated with a living animal, such as audioceptive, electroreception, thermoreception and ecololocation are not likely to be associated with scavenging. For example, bats that rely on echolocation do not lend themselves to discovering immobile carrion.
%
Apart from the basic capacity of these senses to detect carrion, how they function in different environments is also important. In the most simple case the search space is a two dimensional plane (Pawar et al 2012). If the scavenger itself is searching on the plane, such as the case for terrestrial scavengers, the detection range is simply defined by the radius of their sensory organs.
%KH I changed it to make it less confusion with regards to dimension.

%A simplification of terrestrial, vertebrate scavengers in sensory terms is one of them existing in a two-dimensional plane while foraging for carrion directly.
%They can detect carcasses at a range that is defined by the radius of their sensory organs. %, usually the visual and olfactory senses.
For vision in such a scenario, the ability to detect carrion can be seriously restricted. Such species can make up for this though using olfactory sensory abilities which are less restricted.
%KH this kinda jumps ahead about the aerial foraging thing which really dominates the section.
For example, Hyenas make up for this in their ability to smell a rotting carcass 4 km away and to hear the vocalisations of conspecifics at a distance of 10 km \citep{mills1989comparative}. 
% Using the approach of \cite{spiegel2013factors} we estimate a spotted hyena %TG: do we need a (\textit{Crocuta crocuta}) here?
%could resolve a 2 metre target at 1 km distance. 
We can compare this to the energetics approach of \cite{ruxton2004obligate}, who calculated a terrestrial scavenger needs to be able to detect carrion at 500 meters in order to survive, which is clearly within the ability of hyenas.

%%KH I'm not sure of the argument here and I'm kinda finding it hard to grasp the tread of this entire section to be honest.

% While considering prehistoric habitats \cite{ruxton2004obligate} calculated that ``a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m".
Moreover, the senses of many extant (and in all probability extinct) carnivores meet this required distance, making scavenging feasible for terrestrial species \citep{farlow1994speculations,mech2010wolves}.

Species capable of flight have effectively added an extra spatial dimension (i.e. the vertical component) to their sensory environment over land animals.
%I would 
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Certainly, vultures are known to have impressive visual acuity, with one estimate indicating lappet-faced vultures %TG: We need a standardisation for english species names, I'll go for all lower case (similar as what you'd do for "pig" or other english common species names).
 (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
Eagles too are known to have highly developed vision \citep{reymond1985spatial}.
It follows that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts \citep{AR:AR22815}.
%We can contrast this ability with bats, whose visual acuity is famously poor. %TG: Humph! That's not nice! I'm wouldn't go with all the "this is known" thingies here. It really lowers down the quality of our claim and is totally un-necessary (and actually probably wrong: lot's of fruit bats (like flying foxes) do not echolocate and thus have to find their fruits among the flipping tropical folliage using their eyes and noses only)...

@@

%Thus, the bone crushing ability of hyenas and others reveals another useful scavenger trait. %TG: how about the following for a smoother transition:

%KH still not sure about this transition.
Perhaps the most extreme example of utilizing such late carcass stage material is the ability to eat bone, a resource that is typically too hard for many predators to process. 
% Thus, some phenotypic adaptations such as the ability to crush bone reveal another useful scavenger trait.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the estimated 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

@@

Interestingly, such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{ARCM:ARCM12084}.
The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years \citep{dominguez2002hunting}.
%%KH I think this shouldn't be the first place we mention humans as they do appear and fit well in other sections.
A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}.
In some historical hominin-inhabited areas there were a greater number of felids than hyenids.
Again, this is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth, enabling contemporaneous hominins to benefit \citep{pobiner2015new}.
The use of tools and the cooperative nature of hominins meant they could likely get a substantial part of their energetic requirements through scavenging depending on their environment \citep{moleon2014humans}.
%KH should this go in the prey availability section?

In Mesozoic systems some extremely large theropod dinosaurs had a morphology indicative of an ability to process bone \citep[e.g. the robust skull and dentition of \textit{Tyrannosaurus rex}][]{hone2010feeding}.
There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}.
The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating} which would have been powerful enough to break open skeletal material \citep{rayfield2001cranial}.
%KH I would mention that however, our paper shows that such species that would have been able to have such bites were probably too large for this to be an effective stratagy. Then you can segue into the body size thing.
